---
title: "Distribution Models of Visuomotor Adaptation Strategy Development"
author: "Marius 't Hart"
date: "2026-07-06"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

We have done 2 analyses testing how strategy development might be split into discrete styles of cognitive learning. One based on our observations that seemed to point at 3 distinct styles of strategy development which used features, PCA and HCA. The other based on a method pioneered in Tsay et al., 2024, and also used in Ding et al., 2026, which uses dynamic time-warping (DTW) and tSNE. Neither showed clear clusters in the data. This leads us to believe that there is a continuum of strategy development styles. That continuum might be better described as a combination of distributions. That is what we look at here.

# Requirements

First we need to set up the R environment: get the Reach package with helper functions, download the data, and source the code with project-specific functions:

```{r requirement}


# # # # # # # # # # # # # # # # # # # # # # # # #


# Install Reach if not already installed.
# (Marius' package with a bunch of convenience functions.)
if (!require(Reach)) {
  if (!require(remotes)) {
    install.packages(remotes)
  }
  remotes::install_github('thartbm/Reach', ref='main', force=TRUE)
} else {
  # check if the version of Reach if 2026.07.09 or higher:
  if (packageVersion("Reach") < "2026.07.09") {
    if (!require(remotes)) {
      install.packages(remotes)
    }
    remotes::install_github('thartbm/Reach', ref='main', force=TRUE)
  }
}
library(Reach)


# # # # # # # # # # # # # # # # # # # # # # # # #


# load source scripts:

# downloading the data from OSF:
source("R/data.R")

# test all the properties that might describe a distribution of strategy development styles:
source("R/properties.R")


# # # # # # # # # # # # # # # # # # # # # # # # #


# make sure we have the data:

# getData()

```



## Properties of the distribution of strategy development styles

A previous attempt at describing such distributions only considered stepwise strategy development (D'Amario et al., 2024; poster at NCM Panama, 2025). However, especially for laregr rotations, many participants have an exploratory strategy development style. Therefore, we will consider a wider range of properties that might describe the distribution of strategy development styles.

For a stepwise distribution, we need:

1. size of the step (bi-modal)
2. trial the step occurs
3. --
4. sd before the step
5. --
6. sd after the step

For both exploratory (and gradual) strategy development, we need:

1. size of final strategy (bi-modal, same as size of step above)
2. trial at which strategy development starts
3. duration of strategy development (in trials)
4. sd before strategy development starts
5. sd during strategy development
6. sd after strategy stabilizes (maybe two: one for people in each of the modes of the bi-modal strategy size distribution)

So that makes 2 extra parameters (or more likely distributions with multiple parameters). There will be cost in those extra parameters, but we can build both models and test whether or not the extra two distributions pay of sufficiently in additional goodness of fit.

## Data-based properties

We can use the data to estimate the properties of the distribution of strategy development styles. First we extract them for all participants.

> If the file with properties isn't there yet, this could take a few minutes.

```{r extract_parameters}
properties <- getProperties()
head(properties)
```

### All properties

We now have five files with properties that could be used for model building. Just using density functions, and combining all data, this looks like this:

```{r property_density_plot, fig.width=9, fig.height=10}
plotPropertyDistributionsByRotation()
```

Looks like at least some of these properties have a clear distribution, and that some of them vary as a function of rotation size. This can be used to inform distribution-based models. We'll now need to look at each of them in more detail.

### Stepwise components

#### Aiming Step-time and Step-size

First we look at the distributions of step-time and step-size. We can fit a gamma distribution to both of these, and see if there is a relationship between rotation size and the parameters of the gamma distribution. Although it would be better to ignore them if the relationship is not too strong.

```{r steptime_size, fig.width=8, fig.height=8}
plotAimingStepSizeTime()
```

In this case, we will pick a single gamma distribution for step time.

Step size is a bit more complicated. After trimming step-sizes to be larger than -10 and no larger than rotation+10 degrees, we can fit bi-modal distributions the same way as the final strategy size. We fix the means to 0 and half the rotation size. And we get weights that are mostly similar to those of final strategy size. The standard deviations are also highly similar.

#### Aiming Pre- and post-step standard deviations

Now we just need to look at the standard deviations of pre- and post-step intervals. As well as the standard deviations of people without any strategy. Just looking at the initial figure with distributions for all properties, it seems that the `poststep_sd` is very similar to `stable_sd` and `prestep_sd` looks highly similar to `predev_sd` and `devel_sd`. This might also hint at the value of the simpler model, but we will see how this goes.

```{r step_sd_distr, fig.width=9, fig.height=7}
plotAimingStepSD()
```

Separate noise levels for each of the five rotations seems to work better in all cases, so we will use that.

We could draw noise SD for each participant at random from a gamma distribution, but for simplicity we can just use the average of the fitted gamma distribution.

#### Adaptation Step-time and Step-size

First we look at the distributions of step-time and step-size. We can fit a gamma distribution to both of these, and see if there is a relationship between rotation size and the parameters of the gamma distribution. Although it would be better to ignore them if the relationship is not too strong.

```{r steptime_size, fig.width=8, fig.height=8}
plotAdaptationStepSizeTime()
```

Here, we will also pick a single gamma distribution for step time.

Step size is not really uni-modal anymore, so we have compared a normal and gamma distribution.

#### Adaptation Pre- and post-step standard deviations

Now we just need to look at the standard deviations of pre- and post-step intervals. As well as the standard deviations of people without any strategy. Just looking at the initial figure with distributions for all properties, it seems that the `poststep_sd` is very similar to `stable_sd` and `prestep_sd` looks highly similar to `predev_sd` and `devel_sd`. This might also hint at the value of the simpler model, but we will see how this goes.

```{r step_sd_distr, fig.width=7, fig.height=7}
plotAdaptationStepSD()
```
Here again, 5 separate gamma distributions are a better fit than one overall one, for all 3 variables.



### Expanded Model

Here we inspect the Expanded Model Properties

#### Final Strategy

We estimate this bi-modal strategy size using a Gaussian mixture model with 5 parameters (estimated as 6):
- mu 1: mean of the first mode
- mu 2: mean of the second mode
- sigma 1: standard deviation of the first mode
- sigma 2: standard deviation of the second mode
- pi: proportion of the first mode (the second mode is 1-pi)

We notice that the mean of the first mode should be around 0, so we fix it at that value. We've also already seen that the the mean of the strategy learned in each rotation group is roughly half the size of the rotation, so for now we also fix that at 0.5 * rotation. What get's fitted is the standard deviations of the two modes and their weights (pi and 1-pi).

We should develop a super-fit that fits the mean and sigma of the first mode as a constant across rotation sizes and the multiplication factor for the second mode as a constant across rotation sizes. This would also require fitting the relationship between the rotation size and the standard deviation of the second mode, as well as the function of the value of pi depending on rotation size. This seems do-able but we'll leave that for a bit.

First we see if the bi-modal approach matches the data set. (This takes ~20 seconds.)

```{r fit_final_strat_size}
fitFinalStrategyDistributions(properties) # fits stored in a file
```
Now we want to see the fits on top of the data:

```{r final_strat_distr, fig.width=5, fig.height=7}
plotFinalStratDistributions(properties)
```

This works relatively well. Even if we don't believe there are discrete strategy development styles, the bi-modal distribution of final strategy sizes should be replicating the strategy-users vs. non-strategy users from that approach. Here we plot the values of weights (pi and 1-pi) of the modes as a stacked bar graph over rotation size.

```{r mode_weight, fig.width=7, fig.height=5}
plotFinalStratModeWeights()
```

Blue is the weight of the second mode (or 1-pi) and white is the weight of the first mode (or pi).

This indeed looks familiar, and we should be able to do some kind of chi-squared based probability comparison showing equality between this set of probabilities and the ones we got from the DTW- and feature-based approaches.

#### Strategy Development Onset Trial

This distribution could be a Poisson distribution, but in D'Amario et al's data we found it to better fitted as a gamma distribution (makes less sense). We will compare both fits, as well as the fit of a normal distribution. We'll do this for each rotation size separately. But we start with calculating it for all data combined. 

It could be that a simple fit across all data at once is sufficient here. The combined model is 1 single distribution, which would result in 2 parameters. If we fit separate distributions for each of the rotation sizes, that means 5 x 2 = 10 parameters, increasing the penalties for those fits.

```{r devel_onset_fits}
fitOnsetGammaDistributions(properties)
```

Interestingly, the normal distribution fits better (sometimes much better) than the Poisson distribution, but no matter: the gamma distribution fits even better and this matches the previous findings. The overall fit also fits better, so we use one gamma distribution for the onset trial.

Let's plot the fits, both the fits for each rotation size, and the combined gamma distribution:

```{r onset_gamma_plot, fig.width=7, fig.height=7}
plotOnsetGamma(properties)
```

The colored dotted lines indicate the density across participants for each rotation separately, in the same colors we see dots for individual participants, and a thin continuous line indicating the separate gamma fits.

The dashed purple line indicates the combined fit, which is indeed a reasonably good fit for all rotation sizes.

#### Strategy Development Duration

We could model the strategy stabilization trial as a distribution, but it seems cleaner to model the duration of strategy development as a distribution. As long as we ensure that the duration is always a positive number, the stabilization trial will never happen before the onset trial. This is a little harder when modelling the stabilization trial separately from development onset, though of course not impossible.

The duration of the experiment could have lead to a negative correlation between development onset and development duration. Since there is less and less time to develop a strategy the later the development of the strategy starts. We plot that, and test for a correlation. We also look at the distribution of all strategy development durations, and fit several distributions to that.

```{r plot_strat_dev_dur, fig.width=7, fig.height=7}
checkStratDevDuration(properties)
```

First, there is no correlation between the onset trial and the duration of strategy development.

The distribution of strategy development durations (blue area on top) is also not a simple gamma distribution (green line), but a mixture of two distributions (red line). This indeed fits slightly, albeit not significantly (delta AIC=0.3), better than a gamma distribution, even when the gamma distribution has only 2 parameters, whereas the effective number of parameters of the mixture model is 5.

Just to stay somewhat more true to the data, we go with the bi-modal distribution for now, even when this requires special care to make sure we do not draw negative durations from the distribution. There is a subset of people with a strategy that take about 15-20 trials to develop it, and another subset who take almost 80 trials to develop their strategy. It might not matter, but the gamma distribution deviates from this pattern somewhat.

#### Standard Deviations

At this point only the standard deviations of the three phases of strategy development remain. We will look at the standard deviations before, during and after strategy development.

The standard deviation _before_ strategy development is a measure of how much participants explore the space before they start developing a strategy, or when they never explore a strategy. This means that simulated participants in the first mode of the final strategy (with a mean of 0 strategy) will display this level of noise throughout the 120 trials of the task.

The standard deviation _during_ strategy development is a measure of how much participants explore the space while they are developing a strategy. Here we might see effects of _exploratory_ (high SD) and _gradual_ (very low SD) strategy development.

The standard deviation after strategy development is a measure of how consistent participants can report a strategy once it has stabilized. There might be an effect of rotation size in the last level of noise, as aiming closer to the target, may be anchored by the target location (and perhaps its size), and this may help less for larger rotations. Alternatively, only people with low tolerance for errors (very precise people) develop a strategy in lower rotation, and people with more variable response also develop strategies at larger rotations.

```{r sd_pairs, fig.width=6, fig.height=6}
pairs(properties[which(properties$stable_sd < 40),c('predev_sd','devel_sd','stable_sd')], pch=21, col = c("orange","red", "purple", "blue", "turquoise")[(properties$rotation/10)-1])
```
There do not appear to be any clear patterns, but we'll look a little deeper. First, we test if any of the standard deviations are correlated:

```{r}
correlateSDs()
```

Unexpectedly, there is a fairly strong correlation between the pre-development SD and development SD, and also a moderate correlation between the development SD and the stable SD. Despite this, there is no correlation between pre-development SD and stable SD.

This complicates things, but we might have to take this into account. Perhaps we should also test for further relationships between descriptive variables here.

It looks like they are all gamma distributed, but let's plot and test this now:

```{r inspect_sd_dist, fig.width=8, fig.height=8}
plotSDbyRotation(properties)
```

For all subsets of data, the gamma distribution fits better (d AIC > 2) than the normal distribution, except for the 20 degree condition pre-development SD. However the AIC is still lower for the gamma distribution in that case, so we will go with the gamma distribution throughout. Poisson distributions don't work for this data.

Five, rotation specific gamma distributions fit better than a single one (despite the cost for added parameters).

We could model the relationship between rotation size and gamma distribution parameters. This 1) could lead to additional insight, and 2) would reduce the number of parameters. Here we take a look at this:

```{r sd_fit_pars_rotation, fig.width=7, fig.height=7}
plotSDfitPars()
```

It seems that it is possible there some relationship between rotation size and the shape of the distribution from which standard deviations of "noise"/exploration are drawn.

Nevertheless, for now, we will keep it simple and fit five gamma distributions to each of the 3 standard deviations and not worry about correlations. We can always explore if taking these nuances into account substantially improves the model. The goal is first to see if strategy development ramps with potentially increased (exploratory styles) or decreased (incremental styles) standard deviations works better than a step function approach that does not allow for this nuance. If it does, we can always refine the distribution model a bit, but that's not the point.

Perhaps for now the most important point is that the standard deviations before and after strategy development are drawn from more similar distributions than the standard deviations during strategy development. So it seems to make sense to include this, and this is the property that could make the expanded model an improvement over the stepwise model.


# Model building

Now we should line up all the ingredients to start building the two models. Let's get back to the lists of properties of the two models:

Stepwise distribution:

| No. | property              | distribution         |
|-----|-----------------------|----------------------|
| 1   | size of the step      | 5 bi-modal           |
| 2   | trial the step occurs | 1 gamma              |
| 3   | -                     | -                    |
| 4   | sd before the step    | 10 gamma (rot &      |
|     |                       |    w/wo strat)       |
| 5   | -                     | -                    |
| 6   | sd after the step     | 5 gamma              |

For both exploratory (and gradual) strategy development, we need:

| No. | property                | distribution |
|-----|-------------------------|--------------|
| 1   | size of final strategy  | 5 bi-modal   |
| 2   | strat. dev. onset trial | 1 gamma      |
| 3   | strat. dev. duration    | 1 bi-modal   |
| 4   | pre-dev SD              | 5 gamma      |
| 5   | strat. dev. SD          | 5 gamma      |
| 6   | post-dev SD             | 5 gamma      |

## Exponential model?

We could also build an exponential model, which would have these properties:

| No. | property                | distribution |
|-----|-------------------------|--------------|
| 1   | asymptote (final strat) | [5 bi-modal] |
| 2   | -                       | -            |  
| 3   | rate of change          | ??           |
| 4   | SD                      | ??           |  
| 5   | -                       | -            |
| 6   | -                       | -            |

Presumably, the exponential distributions will not work well at all, and either the stepwise or expanded stepwise models will be the best. This remains to be tested though.

Let's have a look at the exponential approach to modeling aiming:

```{r exp_aiming_plot, fig.width=7, fig.height=7}
plotExponentialAiming()
```


And the exponential approach to adaptation:

```{r exp_adapt_plot, fig.width=7, fig.height=7}
plotExponentialAdaptation()
```

We (obviously) need 5 different asymptote levels (although maybe they could be expressed as a proportion of the rotation, and remain constant that way).

The rate of change could be the same across rotations, and a gamma distribution fits a bit better according to distfit, so still need to change this.

The SD is probably a function of rotation size as well, although both the mean and SD of the distribution the noise level is drawn from seem to scale with rotation size, so we could simplify that in the future.


## Adaptation models?

At least for the stepwise and exponential distribution models, we could fit them to the adaptation curves, and see which of the two explains the data best. Presumable, the exponential model will fit best here. This would demonstrate that overall adaptation and explicit strategy development are not the same type of process.

## Stepwise Model

Let's first do the simplest of the two models. It's based on 3 files, each with a set of parameters for each rotation.

Each rotation size gets simulated separately, the goals is to use at least 15k runs. The smallest proportion people in any split is the no-strategy probability in the stepwise approach at 60 degrees with 0.07686916 of participants. If we want at least a 1000 runs in any situations, we'd need 1000/0.07686916 = ~13k runs, but 15k is a little safer.

## Model fitting

The initial model fitting used optim with L-BFGS-B, which essentially returned the starting parameters, but also did not return the lowest negative log-likelihood that was evaluated.

So perhaps this means the starting parameters were already very close to optimal (which they were chosen to be), or the method couldn't handle the model. So we should try out some other methods of optimization. For example, using optimx.

Here are all the optimx methods allowed:

Nelder-Mead
BFGS
CG
L-BFGS-B
nlm
nlminb
spg
ucminf
newuoa
bobyqa
nmkb
hjkb
Rcgmin
Rvmmin

The first 3 definitely do not allow box constraints, and the 4th we used already. For some of the others, they may or may not allow box constraints. If they don't allow box constraints then some of them might allow returning Inf or NA which allows to quasi implements constraints within the function that is to be optimized.

We should first just test which ones work with the box constraints I set up before.